MyTitle

### 2024-11-08 paper/main.tex
updated

\clearpage
\tableofcontents
\clearpage
\chapter{Introduction}

@@

To use Gunther et al.'s formalization, we adopted the c.t.m.\  approach for our proof, as in their independence proof of CH.
Specifically, we assumed the existence of a c.t.m.\  of ZF and constructed a model of ZF + $\neg$AC by forcing.
This model is known as the basic Cohen model, which is a symmetric extension of assumed c.t.m.\ 
Our proof is based on Karagila's lecture note \cite{karagila} and Jech's book \cite{jech_set_theory,jech_AC},
as these resources align well with this approach.

The c.t.m.\  approach is a well-established method for relative consistency proofs in axiomatic set theory.
The relative consistency of $\neg$AC with ZF means that if ZF is consistent, then ZF + $\neg$AC is also consistent.
Assuming the consistency of ZF implies that a model of ZF exists, but strictly speaking, the existence of a c.t.m.\  cannot be derived from this assumption in ZF.
However, as explained in section \ref{subsection:ctm}, we can prove the relative consistency of $\neg$AC with ZF if 
we can construct a model of ZF + $\neg$AC from a c.t.m.\  of ZF.
This kind of reasoning always arises in the c.t.m.\  approach, and ideally, 
we would like to formalize it as well, but this has not been achieved.

@@

\chapter{Preliminaries}
\section{Set-Theoretic Preliminaries}
In this section, we introduce the concepts of set theory used in the formalization of this study.
Our meta-theory is ZF, within which we explore ZF itself. 
Basically, our definition and theorems combine Kunen\cite{kunen2011}, karagila's lecture note\cite{karagila}, and Jech's book\cite{jech_AC}
adapted to the formalized form in Isabelle/ZF.

@@

\begin{dfn}
  The axioms of ZF are the following statements:
  \begin{itemize}
    \item Extensionality: $\forall x \forall y (\forall z (z \in x \leftrightarrow z \in y) \rightarrow x = y)$
    \item Pairing: $\forall x \forall y \exists z \forall w (w \in z \leftrightarrow w = x \lor w = y)$
    \item Union: $\forall x \exists y \forall z (z \in y \leftrightarrow \exists w (z \in w \land w \in x))$
    \item Power set: $\forall x \exists y \forall z (z \in y \leftrightarrow z \subseteq x)$
    \item Infinity: $\exists x (\emptyset \in x \land \forall y (y \in x \rightarrow y \cup \{y\} \in x))$
    \item Regularity: $\forall x(x \neq \emptyset \rightarrow \exists y (y \in x \land y \cap x = \emptyset))$
    \item Infinity: $\forall x (x \neq \emptyset \rightarrow \exists y (y \in x \land \forall z (z \in x \rightarrow z \notin y)))$
    \item Separation: $\forall p \forall x \exists y \forall z (z \in y \leftrightarrow z \in x \land \phi(z, p))$
    \item Replacement: $\forall p (\forall x \forall y \forall z (\phi(x, y, p) \land \phi(x, z, p) \rightarrow y = z) \rightarrow \\ \forall X \exists Y \forall y (y \in Y \leftrightarrow \exists x (x \in X \land \phi(x, y, p))))$
  \end{itemize}
\end{dfn}
Where separation and replacement are axiom schemas, representing infinitely many axioms for each formula $\phi$ with an appropriate arity.
\begin{dfn} 
  \emph{The axiom of choice (AC)} is the following statement: \\
  $\forall x \exists f (\mathrm{"} f \text{ is a function on } x \mathrm{"} \land \forall y (y \in x \rightarrow f(y) \in y))$
\end{dfn}
Where the phrase "f is a function on x" is also considered an abbreviation in the language of set theory.
Theory ZF + AC is denoted by ZFC.
Additionally, we introduce the well-ordering theorem, as we treat AC in this form.

@@

\subsection{The c.t.m.\  Approach}\label{subsection:ctm}

\section{Outline for the Informal Proof}\label{sec:outline}
We outline an informal proof of the relative consistency of $\neg$AC with ZF which we will formalize in the next chapter.
In this proof, the relative consistency is proved by assuming the existence of a c.t.m.\  of ZF and constructing a model of ZF + $\neg$AC by forcing. 
This model is a symmetric extension called the basic Cohen model.

@@

One approach to formalizing this part is the one described in section \ref{subsection:ctm}.
However, to formalize the argument, we would likely need to deal with arbitrary models of ZF, 
not necessarily $\in$-models and prove propositions such as the reflection principle and 
the Mostowski collapse lemma for these models.
Currently, formalized definitions and lemmas related to internal models assume that the ground model is an $\in$-model,
so a large amount of work would be required to formalize this part.

>>>>

\clearpage
\tableofcontents
\clearpage

\chapter{Introduction}

@@

To use Gunther et al.'s formalization, we adopted the c.t.m.\  approach for our proof, as in their independence proof of CH.
Specifically, we assumed the existence of a c.t.m.\  of ZF and constructed a model of ZF+$\neg$AC by forcing.
This model is known as the basic Cohen model, which is a symmetric extension of assumed c.t.m.\ 
Our proof is based on Karagila's lecture note \cite{karagila} and Jech's book \cite{jech_set_theory,jech_AC},
as these resources align well with this approach.

The c.t.m.\  approach is a well-established method for relative consistency proofs in axiomatic set theory.
The relative consistency of $\neg$AC with ZF means that if ZF is consistent, then ZF+$\neg$AC is also consistent.
Assuming the consistency of ZF implies that a model of ZF exists, but strictly speaking, the existence of a c.t.m.\  cannot be derived from this assumption in ZF.
However, as explained in section \ref{subsection:ctm}, we can prove the relative consistency of $\neg$AC with ZF if 
we can construct a model of ZF+$\neg$AC from a c.t.m.\  of ZF.
This kind of reasoning always arises in the c.t.m.\  approach, and ideally, 
we would like to formalize it as well, but this has not been achieved.

@@

\chapter{Preliminaries}

\section{Set-Theoretic Preliminaries}
In this section, we introduce the concepts of set theory used in the formalization of this study.
Our meta-theory is ZF, within which we explore ZF itself. 
Basically, our definition and theorems combine Kunen\cite{kunen2011}, karagila's lecture note\cite{karagila}, and Jech's book\cite{jech_AC}
adapted to the formalized form in Isabelle/ZF.

@@

\begin{dfn}
  The axioms of ZF are the following statements:
  \begin{itemize}
    \item Extensionality: $\forall x \forall y (\forall z (z \in x \leftrightarrow z \in y) \rightarrow x = y)$
    \item Pairing: $\forall x \forall y \exists z \forall w (w \in z \leftrightarrow w = x \lor w = y)$
    \item Union: $\forall x \exists y \forall z (z \in y \leftrightarrow \exists w (z \in w \land w \in x))$
    \item Power set: $\forall x \exists y \forall z (z \in y \leftrightarrow z \subseteq x)$
    \item Infinity: $\exists x (\emptyset \in x \land \forall y (y \in x \rightarrow y \cup \{y\} \in x))$
    \item Regularity: $\forall x(x \neq \emptyset \rightarrow \exists y (y \in x \land y \cap x = \emptyset))$
    \item Infinity: $\forall x (x \neq \emptyset \rightarrow \exists y (y \in x \land \forall z (z \in x \rightarrow z \notin y)))$
    \item Separation: $\forall p \forall x \exists y \forall z (z \in y \leftrightarrow z \in x \land \phi(z, p))$
    \item Replacement: $\forall p (\forall x \forall y \forall z (\phi(x, y, p) \land \phi(x, z, p) \rightarrow y = z) \rightarrow \\ \forall X \exists Y \forall y (y \in Y \leftrightarrow \exists x (x \in X \land \phi(x, y, p))))$
  \end{itemize}
\end{dfn}
Where separation and replacement are axiom schemas, representing infinitely many axioms for each formula $\phi$ with an appropriate arity.
\begin{dfn} 
  \emph{The axiom of choice (AC)} is the following statement: \\
  $\forall x \exists f (\mathrm{"} f \text{ is a function on } x \mathrm{"} \land \forall y (y \in x \rightarrow f(y) \in y))$
\end{dfn}
Where the phrase "f is a function on x" is also considered an abbreviation in the language of set theory.
Theory ZF+AC is denoted by ZFC.
Additionally, we introduce the well-ordering theorem, as we treat AC in this form.

@@

\subsection{The c.t.m.\  Approach}\label{subsection:ctm}
In this section, we explain the validity of the c.t.m.\  approach, 
which is a method for proving the relative consistency and independence in axiomatic set theory.

The relative consistency of $\neg$AC with ZF means that if ZF is consistent, then ZF+$\neg$AC is also consistent.
When ZF is consistent, we can prove, by the completeness theorem, that ZF has a model.
However, the existence of a c.t.m.\  of ZF cannot be derived from the consistency of ZF.
The statement that ZF has a c.t.m.\  is stronger than the consistency of ZF.
Nonetheless, by assuming the existence of a c.t.m.\  of ZF and then constructing a model of ZF+$\neg$AC by forcing, 
we can prove the relative consistency of $\neg$AC with ZF.
To explain this, we introduce some concepts,
following Jech \cite{jech_set_theory}, chapter 6 and 12.

\begin{prop}[Reflection Principle]
Let $\varphi(x_1, \ldots, x_n)$ be a formula. There is a transitive set $M$ such that 
$$\varphi^M(x_1, \ldots, x_n) \leftrightarrow \varphi(x_1, \ldots, x_n)$$
for every $x_1, \ldots, x_n \in M$.
\end{prop}
Where $\varphi^M$ denotes the relativization of $\varphi$ to $M$.
In other words, $\varphi^M$ is the formula obtained 
by replacing every quantifier $\forall x$ by $\forall x \in M$ and $\exists x$ by $\exists x \in M$.

\begin{dfn}
  Let $E$ be a binary relation on a set $P$.
  \begin{itemize}
    \item 
  We say thet $E$ is \emph{well-founded} if every non-empty subset of $P$ has an $E$-minimal element.
  \item
  We say that $E$ is extensional if:
  $$\mathrm{ext}_E(x) \ne \mathrm{ext}_E(y)$$
  whenever $x$ and $y$ are distinct elements of $P$, 
  where 
  $$\mathrm{ext}_E(x) = \{ z \in P \mid z\ E\ x \}$$ 
\end{itemize}
\end{dfn}

\begin{thm}[Mostowski's Collapsing Theorem]
  If $E$ is a well-founded and extensional relation on a set $P$, 
  then there is a transitive set $M$ and an isomorphism $\pi$ between $(P, E)$ and $(M, \in)$.
\end{thm}

Additionally, we can prove weaker form of the Löwenheim-Skolem theorem without the assumption of AC
\footnote{Following Henkin's proof of the completeness theorem, we can obtain a countable model of the theory without using AC.}.
\begin{thm}\label{thm:lowenheim_skolem}
  If there exists a model of a theory of a countable language, then there exists a countable model of the theory.
\end{thm}

Using these, we can prove the following proposition:
\begin{prop}
  Let $\Gamma$ be a finite fragment of ZF axioms,
  then there exists a c.t.m.\ of $\Gamma$.
\end{prop}
\begin{proof}
  Let $\Gamma = \{ \varphi_1, \ldots, \varphi_n \}$ and $\psi = \varphi_1 \land \ldots \land \varphi_n$.
  By reflection principle, there exists a set $M$ such that $\psi^M \leftrightarrow \psi$.
  Then, $(M, \in \mid_{M \times M})$ satisfies $\psi$. 
  In other words, $(M, \in \mid_{M \times M})$ is a model of $\Gamma$.
  By Theorem \ref{thm:lowenheim_skolem}, there exists a countable model $(N, R)$ of $\Gamma$.
  Since $(N, R)$ satisfies the extensionality axiom, $R$ is extensional.
  Also, 
\end{proof}

\section{Outline for the Informal Proof}\label{sec:outline}
We outline an informal proof of the relative consistency of $\neg$AC with ZF which we will formalize in the next chapter.
In this proof, the relative consistency is proved by assuming the existence of a c.t.m.\  of ZF and constructing a model of ZF+$\neg$AC by forcing. 
This model is a symmetric extension called the basic Cohen model.

@@

One approach to formalizing this part is the one described in section \ref{subsection:ctm}.
However, to formalize the argument, we would likely need to deal with arbitrary models of ZF, 
not necessarily $\in$-models
\footnote{An $\in$-model is a model $(M, R)$ of set theory such that $R = \ \in \mid_{(M \times M)}$, i.e. $R$ is actual membership.}, 
and prove propositions such as the reflection principle and 
the Mostowski collapse lemma for these models.
Currently, formalized definitions and lemmas related to internal models assume that the ground model is an $\in$-model,
so a large amount of work would be required to formalize this part.
